The effect of linking assumptions and number of response options on inferred scalar implicature rate

### 2018-05-08 writeup.Rmd
minor edits and comments on gd

To explain these data on the traditional view, we are forced to say that a) not all participants calculated the implicature; or that b) some participants who calculated the implicature did not choose the anticipated (i.e., "Wrong") response due to some other cognitive process which overrode the 'correct' implicature behavior; or some mixture of (a) and (b). We might similarly posit that one or both of these factors underlie the variation in the ternary, quaternary, and quinary conditions. However, without an understanding of how to quantitatively specify the link between implicature calculation and its behavioral expression, the best we can hope for on this approach is an analysis which predicts general qualitative patterns in the data (e.g. a prediction of relatively more "Right" responses than "Wrong" responses in a given trial of our binary truth-value judgment task, or a prediction of a rise in the rate of response of "Right"/"Wrong" between two experimental conditions, given some contextual manipulation). **XXX Judith, I was thinking of adding the sentence to end this paragraph: "However, we should stress that to the best of our knowledge, even a qualitative analysis of this kind of variation in behavior on sentence verification tasks - much less the effect of the number of response choices on that behavior - is largely underdeveloped scalar implicature literature." But it might be way too strong a claim? **

**XXX Brandon, can you be more specific about what you mean by 'tracing a general qualitative pattern'? XXX Judith: I've tried to offer some examples of what I mean by this, see above.**

We contrast the above view of implicature and its behavioral reflexes with an alternative linking hypothesis. Recent developments in the field of probabilistic pragmatics have demonstrated that pragmatic production and comprehension can be captured within the Rational Speech Act (RSA) framework, which treats listeners and speakers as agents that reason about each other and choose interpretations and utterances, respectively, that soft-maximize utility [@Frank2012; @Goodman2013; @Goodman2016; @Franke2016; @Qing2015; @Degen2013; @DegenTG2015; @Bergen2016; @Kao2014; @Scontras2017]. **XXX Judith: re: your Slack comments, in the following few sentences I've provided a summary of the intuition behind RSA (going off of the discussion in Goodman & Stuhlmueller 2013)** Much in the spirit of Gricean approaches to pragmatic competence, the RSA framework takes as its point of departure the idea that individuals are rational, goal-oriented communicative agents, who in turn assume that their interlocutors similarly behave according to general principles of cooperativity in communication. Just as in more traditional Gricean pragmatics, pragmatic inference and pragmatically-cooperative language production in the RSA framework are, at their core, the product of counterfactual reasoning about alternative utterances that one might produce (but does not, in the interest of cooperativity). However, the RSA framework explicitly and quantitatively models coooperative interlocutors as agents whose language production and comprehension is a function of Bayesian probalistic inference regarding other interlocutors' expected behavior in a discourse context.

@@

Thus, our alternative linking hypothesis contrasts with the traditional view in that it is rooted in a quantitative formalization of pragmatic competence which provides us a continuous measure of pragmatic reasoning. In the RSA framework, individuals never categorically draw (or fail to draw) pragmatic inferences about the utterances they hear. For example, exclusivity readings of disjunction are represented in RSA as relatively lower posterior conditional probability of a conjunctive meaning on the $P_L$ distribution given an utterance of "or", compared the prior probability of that meaning. Thus, absent auxiliary assumptions about what exactly would constitute 'implicature', it is not even possible to talk about rate of implicature calculation in the RSA framework. The upshot, as we show below, is that this view of pragmatic competence does allow us to talk explicitly and quantitatively about rates of observed behavior in sentence verification tasks.

@@

A main desideratum of a behavioral linking hypothesis given the RSA view of pragmatic competence is to transform continuous probability distributions into categorical outputs (e.g. responses of “Right”/”Wrong” in the case of the binary condition of our experiment). For a given utterance $u$ and an intended communicated meaning $s$, $S_1$(u | s) outputs a conditional probability of $u$ given $s$. For example, in the binary condition of our experiment where a participant evaluated "There is a cat or a dog" when there were both animals on the card, the participant has access to the mental representation of $S_1$ and hence to the $S_1$ conditional probability of producing the utterance "cat or dog"" given a dog and cat card: $S_1$("cat or dog" | cat and dog). According to the linking hypothesis advanced here, the participant provides a particular response to $u$ if the RSA speaker probability of $u$ lies within a particular probability interval, given an observed state of the world (i.e. the configuration of animals on the card in our experiment). We model a responder, $R$, who in the binary condition responds “Right” to an utterance $u$ in world $s$ just in case $S_1(u | s)$ exceeds some probability threshold $\theta$:

@@

= “Right” iff $S_1$(u | s) $>$ $\theta$

= “Wrong” otherwise

**XXX The following paragraph is mostly new:** Because $\theta$ is a free parameter in our model, we have a straightforward understanding of between-subject response variation on a given trial of our experiment according to this linking hypothesis. That is, we can attribute this variation to individual differences vis a vis the value of the probability threshold against which utterances are evaluated in a given world state. Moreover, the model of a responder in the binary condition is extended intuitively to the condition where participants had three response options. In this case, we allow for two probability thresholds: $\theta_1$, the minimum standard for an utterance in a given world state to count as "Right", and $\theta_2$, the minimum standard for "Neither". Thus, in the ternary condition, R(u, s, $\theta_1$ , $\theta_2$) is “Right” iff $S_1$(u | s) > $\theta_1$ and “Neither” iff $\theta_1$ > $S_1$(u | s) > $\theta_2$.
To fully generalize the model to our five experimental conditions, we say that $R$ takes as its input an utterance $u$, a world state $s$, and a number of threshold variables dependent on a variable $c$, corresponding to the experimental condition in which the participant finds herself (e.g. the range of possible responses available to $R$).

@@

= “Right” iff $S_1$(u | s) $>$ $\theta_1$

= “Neither” iff $\theta_1$ $>$ $S_1$(u | s) $>$ $\theta_2$

= “Wrong” otherwise

Given c = “quatenary”

@@

= “Right” iff $S_1$(u | s) $>$ $\theta_1$

= “Kinda Right” iff $\theta_1$ $>$ $S_1$(u | s) $>$ $\theta_2$

= “Kinda Wrong” iff $\theta_2$ $>$ $S_1$(u | s) $>$ $\theta_3$

= “Wrong” otherwise

@@

= “Right” iff $S_1$(u | s) $>$ $\theta_1$

=“Kinda Right” iff $\theta_1$ $>$ $S_1$(u | s) $>$ $\theta_2$

= “Neither” iff $\theta_2$ $>$ $S_1$(u | s) $>$ $\theta_3$

= “Kinda Wrong” iff $\theta_3$ $>$ $S_1$(u | s) $>$ $\theta_4$

= “Wrong” otherwise

The above analysis is a proof of concept for the following idea: by relaxing the assumptions of the traditional view of scalar implicature (namely, that scalar implicatures either are or are not calculated, and that behavior on sentence verification tasks directly reflects this binary interpretation process), we can propose quantitative models of the variation in behavior we observe in experimental settings.

**XXX The following is entirely new:** However, this way of understanding behaivor in truth value jugdment tasks also raises new analytic and empirical questions which we must leave to future work. For instance, given the experimental results reported here, what is the range of plausible values for, e.g., the single probability threshold $\theta$, which separates "Right" from "Wrong" utterances in the binary condition? Bayesian statistical methods provide us the means for estimating the value of $\theta$ in our RSA-based model, by allowing us to recover a posterior distribution of possible values of $\theta$ given an assumption of a uniform prior and our empirically-observed data, i.e. the actual rate of response of "Right" On each trial of the binary condition of the experiment. We may also ask, for example, whether the speaker probability threshold for an utterance being "Right" is relatively constant across conditions or changes systematically with the introduction of more response options. Again, Bayesian data analysis allows us to explore this question by investigating the extent to which the inferred posterior distribution of, e.g., possible values of the threshold for "Right" in the binary condition of the experiment differs from the inferred posterior distribution of possible values of the threshold for "Right" in the ternary, quatenary, and quinary conditions.

@@

>>>>

To explain these data on the traditional view, we are forced to say that a) not all participants calculated the implicature; or that b) some participants who calculated the implicature did not choose the anticipated (i.e., "wrong") response due to some other cognitive process which overrode the 'correct' implicature behavior; or some mixture of (a) and (b). We might similarly posit that one or both of these factors underlie the variation in the ternary, quaternary, and quinary conditions. However, without an understanding of how to quantitatively specify the link between implicature calculation and its behavioral expression, the best we can hope for on this approach is an analysis which predicts general qualitative patterns in the data (e.g. a prediction of relatively more "right" responses than "wrong" responses in a given trial of our binary truth-value judgment task, or a prediction of a rise in the rate of response of "right"/"wrong" between two experimental conditions, given some contextual manipulation). However, we should stress that to the best of our knowledge, even a qualitative analysis of this kind of variation in behavior on sentence verification tasks - much less the effect of the number of response choices on that behavior - is largely underdeveloped scalar implicature literature.

We contrast the above view of implicature and its behavioral reflexes with an alternative linking hypothesis. Recent developments in the field of probabilistic pragmatics have demonstrated that pragmatic production and comprehension can be captured within the Rational Speech Act (RSA) framework  [@Frank2012; @Goodman2013; @Goodman2016; @Franke2016; @Qing2015; @Degen2013; @DegenTG2015; @Bergen2016; @Kao2014; @Scontras2017]. Much in the spirit of Gricean approaches to pragmatic competence, the RSA framework takes as its point of departure the idea that individuals are rational, goal-oriented communicative agents, who in turn assume that their interlocutors similarly behave according to general principles of cooperativity in communication. Just as in more traditional Gricean pragmatics, pragmatic inference and pragmatically-cooperative language production in the RSA framework are, at their core, the product of counterfactual reasoning about alternative utterances that one might produce (but does not, in the interest of cooperativity). However, the RSA framework explicitly and quantitatively models cooperative interlocutors as agents whose language production and comprehension is a function of Bayesian probabilistic inference regarding other interlocutors' expected behavior in a discourse context.

@@

This view contrasts with the traditional view in that it is rooted in a quantitative formalization of pragmatic competence which provides us a continuous measure of pragmatic reasoning. In the RSA framework, individuals never categorically draw (or fail to draw) pragmatic inferences about the utterances they hear. For example, exclusivity readings of disjunction are represented in RSA as relatively lower posterior conditional probability of a conjunctive meaning on the $P_L$ distribution given an utterance of "or", compared to the prior probability of that meaning. Thus, absent auxiliary assumptions about what exactly would constitute 'implicature', it is not even possible to talk about rate of implicature calculation in the RSA framework. The upshot, as we show below, is that this view of pragmatic competence does allow us to talk explicitly and quantitatively about rates of observed behavior in sentence verification tasks.

@@

A main desideratum of a behavioral linking hypothesis given the RSA view of pragmatic competence is to transform continuous probability distributions into categorical outputs (e.g. responses of “right”/”wrong” in the case of the binary condition of our experiment). For a given utterance $u$ and an intended communicated meaning $s$, $S_1$(u | s) outputs a conditional probability of $u$ given $s$. For example, in the binary condition of our experiment where a participant evaluated "There is a cat or a dog" when there were both animals on the card, the participant has access to the mental representation of $S_1$ and hence to the $S_1$ conditional probability of producing the utterance "cat or dog"" given a dog and cat card: $S_1$("cat or dog" | cat and dog). According to the linking hypothesis advanced here, the participant provides a particular response to $u$ if the RSA speaker probability of $u$ lies within a particular probability interval, given an observed state of the world (i.e. the configuration of animals on the card in our experiment). We model a responder, $R$, who in the binary condition responds “right” to an utterance $u$ in world $s$ just in case $S_1(u | s)$ exceeds some probability threshold $\theta$:

@@

= “right” iff $S_1$(u | s) $>$ $\theta$

= “wrong” otherwise

Because $\theta$ is a free parameter in our model, we have a straightforward understanding of between-subject response variation on a given trial of our experiment according to this linking hypothesis. **JD: Hold on, Brandon, what's the link between being a free parameter hand having an understanding of between-subject response variation?** That is, we can attribute this variation to individual differences vis a vis the value of the probability threshold against which utterances are evaluated in a given world state. Moreover, the model of a responder in the binary condition is extended intuitively to the condition where participants had three response options. In this case, we allow for two probability thresholds: $\theta_1$, the minimum standard for an utterance in a given world state to count as "right", and $\theta_2$, the minimum standard for "neither". Thus, in the ternary condition, R(u, s, $\theta_1$ , $\theta_2$) is “right” iff $S_1$(u | s) > $\theta_1$ and “neither” iff $\theta_1$ > $S_1$(u | s) > $\theta_2$.
To fully generalize the model to our five experimental conditions, we say that $R$ takes as its input an utterance $u$, a world state $s$, and a number of threshold variables dependent on a variable $c$, corresponding to the experimental condition in which the participant finds themself (e.g. the range of possible responses available to $R$).

@@

= “right” iff $S_1$(u | s) $>$ $\theta_1$

= “neither” iff $\theta_1$ $>$ $S_1$(u | s) $>$ $\theta_2$

= “wrong” otherwise

Given c = “quaternary”

@@

= “right” iff $S_1$(u | s) $>$ $\theta_1$

= “kinda right” iff $\theta_1$ $>$ $S_1$(u | s) $>$ $\theta_2$

= “kinda wrong” iff $\theta_2$ $>$ $S_1$(u | s) $>$ $\theta_3$

= “wrong” otherwise

@@

= “right” iff $S_1$(u | s) $>$ $\theta_1$

=“kinda right” iff $\theta_1$ $>$ $S_1$(u | s) $>$ $\theta_2$

= “neither” iff $\theta_2$ $>$ $S_1$(u | s) $>$ $\theta_3$

= “kinda wrong” iff $\theta_3$ $>$ $S_1$(u | s) $>$ $\theta_4$

= “wrong” otherwise

The above analysis is a proof of concept for the following idea: by relaxing the assumptions of the traditional view of scalar implicature (namely, that scalar implicatures either are or are not calculated, and that behavior on sentence verification tasks directly reflects this binary interpretation process), we can propose quantitative models of the variation in behavior we observe in experimental settings. **JD: Brandon, I feel what's still missing here is a demonstration that this is a _reasonable_ linking function. Maybe we can add the point that with this linking function, there's nothing inherently special about the implicature conditions -- the same linking function can work for any of the eight conditions in our experiment, because they're all the result of a participant having to make a behavioral choice in the face of an underlying speaker probability.**

However, this way of understanding behavior in truth value judgment tasks also raises new analytic and empirical questions which we must leave to future work. For instance, given the experimental results reported here, what is the range of plausible values for, e.g., the single probability threshold $\theta$, which separates "right" from "wrong" utterances in the binary condition? Bayesian statistical methods provide us the means for estimating the value of $\theta$ in our RSA-based model, by allowing us to recover a posterior distribution of possible values of $\theta$ given an assumption of a uniform prior and our empirically-observed data, i.e., the actual rate of response of "right" on each trial of the binary condition of the experiment. We may also ask, for example, whether the speaker probability threshold for an utterance being "right" is relatively constant across conditions or changes systematically with the introduction of more response options. Again, Bayesian data analysis allows us to explore this question by investigating the extent to which the inferred posterior distribution of, e.g., possible values of the threshold for "right" in the binary condition of the experiment differs from the inferred posterior distribution of possible values of the threshold for "right" in the ternary, quatenary, and quinary conditions.

@@

**JD: We need a final concluding paragraph that reiterates the main points.**
